Mendelian randomization for causal inference in Genome-wide association studies

### 2022-02-04 mr-review.Rmd
added eqs

Three assumptions are central to MR. Satisfying them is necessary to infer 
causal effects in an observational MR study. Careful evaluation of the assumptions is needed. Briefly, 
the first assumption states that the exposure is associated with the outcome. The second requires 
independence, conditional on the exposure and all measured and unmeasured confounders, between the 
genetic marker and the outcome. The last of the three assumptions states that the genetic marker is 
independent of all (measured and unmeasured) confounders of the relationship between exposure and 
outcome. We next discuss these in more detail and consider their evaluation in MR studies.

@@

The third assumption, sometimes called the , ensures that .

When the three MR assumptions hold, one may use instrumental variable statistical methods, 
with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the 
causal effect of the exposure on the outcome.

@@

Two-stage methods use two regressions. In the first stage, the exposure is regressed on
the instrument to obtain fitted values. In the second stage, the outcome is regressed on the fitted values for the 
exposure, which result from the first regression. Note that the uncertainty in the fitted values of 
the first regression is not considered when performing the second regression.

A two-stage statistical model involves two regression models [@burgess_review_2017]. For a continuous 
outcome, one may choose to perform "two-stage least squares", which involves two linear regressions. 
First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes). The resulting 
fitted values for the exposure variables are the independent variable in the second linear regression, 
where the outcome variable is the dependent variable. The causal effect estimate, then, is the 
coefficient obtained from the second regression analysis.

@@

```{r, dag1, fig.cap = "Scenario where the genetic variant affects an intermediate variable W on the pathway to the exposure X. This violates the exclusion restriction assumption."}
g1 <- dagitty::dagitty( "dag {
	G -> W -> X -> Y
	W -> Y
}")
plot(dagitty::graphLayout(g1))
```

Recognition of the many ways that the MR assumptions may be violated has both inspired methods advances that 
enable assumption relaxations under certain conditions and motivated use of sensitivity analysis to quantify the impact of possible assumption violations.

>>>>

Three assumptions are central to MR. Satisfying them is necessary to infer 
causal effects in an observational MR study. In practice, careful evaluation of the assumptions is needed. 
The first assumption states that the exposure is associated with the outcome. The second requires 
independence, conditional on the exposure and all measured and unmeasured confounders, between the 
genetic marker and the outcome. The last of the three assumptions states that the genetic marker is 
independent of all (measured and unmeasured) confounders of the relationship between exposure and 
outcome. We next discuss these in more detail and consider their evaluation in MR studies [@vanderweele_methodological_2014].

@@

The third assumption ensures that the genetic marker is
independent of all confounders of the exposure-outcome relationship.

When the three MR assumptions hold, one may use instrumental variable statistical methods, 
with the genetic marker data, the exposure data, and the outcome data, to estimate and to test the 
causal effect of the exposure on the outcome [@didelez_mendelian_2007;@lawlor_mendelian_2008;@palmer_adjusting_2008].

To understand the origin of the term "Mendelian randomization", it is helpful to 
recall Mendel's inheritance laws. Under the assumption that alleles segregate 
randomly from parent to offspring, the offspring genotypes are unlikely to be 
associated with confounders of the exposure-outcome relationship. Additionally,
reverse causation is ruled out by the observation that germ-line genotypes are 
fixed at conception and, thus, precede other observed variables.

@@

A two-stage statistical model involves two regression models [@burgess_review_2017]. For a continuous 
outcome, one may choose to perform "two-stage least squares", which involves two linear regressions. 
First, the exposure variable is regressed on the instrument (*i.e.*, the SNP genotypes) (Equation \@ref(eq:lm1)). The resulting 
fitted values for the exposure variable are the independent variable in the second linear regression, 
where the outcome variable is the dependent variable. The causal effect estimate, then, is the 
coefficient obtained from the second regression analysis.

\begin{equation}
  x_i = \alpha_0 + \alpha_1 z_i + \epsilon_i
  (\#eq:lm1)
\end{equation}

where the $\epsilon_i$ are assumed to be iid normal random variables with mean zero.

\begin{equation}
  y_i = \beta_0 + \beta_1 \hat x_i + \tau_i
  (\#eq:lm2)
\end{equation}

where the $\tau_i$ are assumed to be iid normal random variables with mean zero and $\hat x_i$ denotes the fitted value for $x_i$ obtained from Equation \@ref(eq:lm1).

Note that the uncertainty in the fitted values from 
the first regression is not considered when performing the second regression.
For this reason, the variance of the coefficient estimator, $Var(\hat \beta_1)$ is incorrect in two-stage calculations. This observation led researchers to develop likelihood-based MR methods.

@@

```{r, dag1, fig.cap = "Scenario where the genetic variant G affects an intermediate variable W on the pathway to the exposure X. This violates the exclusion restriction assumption."}
g1 <- dagitty::dagitty( "dag {
	G -> W -> X -> Y
	W -> Y
}")
plot(dagitty::graphLayout(g1))
```

Recognition of the many ways that the MR assumptions may be violated has inspired methods advances that 
enable assumption relaxations under certain conditions and has motivated use of sensitivity analysis to quantify the impact of possible assumption violations.
